Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-01-13 paper.Rnw
XVal updated

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
%  - fix figure 2 schematic language for the experiments (not 2a 2b). 
%  - ADNI-1 or ADNI1, standardize
%  - Don't use "Experiment X", instead use some short form description, and then
%  reference (Experiment X).

@@

The gold standard for neuroanatomical segmentation is manual delineation by an
expert human rater.  However, with the availability of increasingly large MRI
datasets, the time and expertise required for manual segmentation becomes
prohibitive \citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. This
effort is complicated by the fact that there is significant variation between
segmentation protocols with respect to specific anatomical boundaries of the
hippocampus \citep{Geuze2004} and this has led to efforts to create an unified
hippocampal segmentation protocol \citep{Jack2011,Boccardi2013,Boccardi2013a}.
In addition, there is controversy over the appropriate manual segmentation
protocol to use in a particular imaging study \citep{Nestor2012}. Thus, a
segmentation algorithm that can easily adapt to different manual segmentation
definitions would be of significant benefit to the neuroimaging community.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                            Experiments                              %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experiments}

The following section describes experiments conducted to assess the
segmentation quality of the \mb{} algorithm. The first two experiments assess the
validity of \mb{} using cross-validation designs. Experiment 1 investigates the
accuracy of \mb{} whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. The last two experiments assess the
validity of the \mb{} algorithm when applied to different diseases: first episode
schizophrenia (Experiment 3), and Alzheimer's disease (Experiment 4).
Additionally, in Experiment 4, we compare \mb{} segmentations with those of
well-known automated methods and assessed segmentation bias.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 1: Whole Hippocampus Cross-Validation            %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 1: Whole Hippocampus Cross-Validation}

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using images
from the ADNI dataset \citep{Jack2008}, and manual whole hippocampus
segmentations following the protocol laid out in \citep{Pruessner2000}.  This
form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb{} (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

In the Supplementary Materials we have replicated this experiment using the
\snt{} semi-automated labels included with the ADNI dataset.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2: Hippocampal Subfield Cross-Validation}

The previous experiment assesses \mb{} peformance on whole hippocampus
segmentation. In this experiment, we evaluate \mb{} hippocampal subfield
segmentation of standard 3T T1-weighted images at $0.9m^3$-isotropic voxels.  We
use a modified leave-one-out cross-validation (LOOCV) design.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Results}

\begin{figure}
<<fig:WAxval-kappa>>=
wa.n.sim = read.csv('data/winterburn.xval.similarity.csv')
wa.n.sim$label = as.factor(wa.n.sim$label)
levels(wa.n.sim$label) <-list(
  "CA1"=1, "CA1"=101, 
  "CA2/CA3"=5,"CA2/CA3"=105, 
  "CA4/DG"=4,"CA4/DG"=104, 
  "Subiculum"=2,"Subiculum"=102, 
  "SR/SL/SM"=6,"SR/SL/SM"=106)

# kappa vs. region
ggplot(wa.n.sim, aes(x=as.factor(label),y=k)) + 
  geom_boxplot(fill='lightblue') + 
  xlab("Subregion") + 
  ylab("Dice's Similarity Coefficient (DSC)")

CA1   = subset(wa.n.sim,label=='CA1')$k
CA23  = subset(wa.n.sim,label=='CA2/CA3')$k
CA4DG = subset(wa.n.sim,label=='CA4/DG')$k
Sub   = subset(wa.n.sim,label=='Subiculum')$k
SRLM  = subset(wa.n.sim,label=='SR/SL/SM')$k
@
  \caption{Similarity of \mb{} segmentation of subregions and the resampled
  Winterburn atlas segmentations at $0.9mm^3$ voxel resolution, over all folds of a
  leave-one-out cross validation. Overlap score for each hemisphere is measured
  separately. \label{fig:WAxval-kappa}}
\end{figure}

Figure \ref{fig:WAval-kappa} shows the overlap similarity scores between the
\mb{} segmentations and the resampled Winterburn atlases for each hippocampal
subregion across all subjects and folds of the validation. Mean and standard
deviation DSC scores of the subregions are shown in Table
\ref{tab:WAxval-perturb-kappa}, along with DSC scores for the resampled atlas
segmentations when perturbed slightly and compared to the originals.  We find
that the CA4/DG subregion shows the highest mean DSC score of
\Sexpr{mean(CA4DG)} $\pm$\Sexpr{sd(CA4DG)}, followed by the Subiculum and CA1
subregions having scores of \Sexpr{mean(CA1)} $\pm$\Sexpr{sd(CA1)} and
\Sexpr{mean(Sub)} $\pm$\Sexpr{sd(Sub)}, respectively.  Both the CA4/DG and
molecular regions score below 0.5.  These scores may seem low but not when taken
in context and compared to existing (semi-)automated methods (see Discussion).

\begin{figure}
<<fig:WAxval-Bland-Altman>>=
wa.n.vol = read.csv('data/winterburn.xval.volumes.csv')
winterburn = read.csv('data/winterburn.volumes.csv')
subfield_labels <-list(
  "CA1"="X1", "CA1"="X101", 
  "CA2/CA3"="X5","CA2/CA3"="X105", 
  "CA4/DG"="X4","CA4/DG"="X104",
  "Subiculum"="X2","Subiculum"="X102",
  "SR/SL/SM"="X6","SR/SL/SM"="X106")
hemispheres <- list(
  "Right"="X1","Left"="X101", 
  "Right"="X5","Left"="X105", 
  "Right"="X4","Left"="X104",
  "Right"="X2","Left"="X102",
  "Right"="X6","Left"="X106")

# mean per subfield/hemisphere across all 4 validation folds
wa.n.mean.vols = ddply(wa.n.all.vols, c("variable", "subject", "res"), function(df) {
  data.frame(mb = mean(df$value),
             high  = mean(df$'0.3'),
             low   = mean(df$'0.9'))
})
wa.n.mean.vols$diff = wa.n.mean.vols$low - wa.n.mean.vols$mb
wa.n.mean.vols$mean = (wa.n.mean.vols$low + wa.n.mean.vols$mb)/2

wa.n.mean.vols$hemisphere <- wa.n.mean.vols$variable
levels(wa.n.all.vols$variable)   <- subfield_labels
levels(wa.n.mean.vols$variable)  <- subfield_labels
levels(wa.n.all.vols$hemisphere) <- hemispheres

wa.n.xval.bland.cor = ddply(wa.n.mean.vols,c("variable"),function(df) {
  cor.t = cor.test(df$low,df$mb)
  data.frame(
    pearson = sprintf("r=%0.2f, t=%0.2f, p < %0.2f", 
                cor.t$estimate, cor.t$statistic, cor.t$p.value)
  )
})

ggplot(wa.n.mean.vols, aes(x=mean, y=diff)) + 
  facet_wrap(~variable, ncol=3, scales='free_x') + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=2, aes(colour=hemisphere, shape=subject)) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=wa.n.xval.bland.limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", 
             data = wa.n.xval.bland.limits, hjust=1.1,vjust=-0.8, size=2.5) + 
  geom_text(aes(y = Inf, x=-Inf, label=pearson), colour="black", 
             data = wa.n.xval.bland.cor, hjust=-0.1,vjust=2.5, size=2.5) + 
  xlab(expression('Mean of resampled and MAGeT-Brain subregion volume ' (mm^3))) + 
  ylab(expression('Difference of manual and MAGeT-Brain subregion volume ' (mm^3))) + 
  #scale_x_continuous(limits=c(0,1000)) + 
  #scale_color_discrete(name="") + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@
  \caption{\textbf{Bland-Altman plots comparing unilateral hippocampal subregion volume
across all folds of Winterburn atlas cross validation.} The overall mean
difference in volume, and limits of agreement ($\pm 1.96SD$) are shown by dashed
horizontal lines. Linear fit lines are shown. Note, points below the mean
difference indication overestimation of the volume with respect to the resampled
volume, and vice versa. \label{fig:WAxval-Bland-Altman}}
\end{figure}

We find that in each subfield, \mb{} shows a proportional bias towards
underestimating volumes in all subregions (Figure
\ref{fig:WAxval-Bland-Altman}).  Prominent fixed biases in the volumes of the
CA4/DG and SR/SL/SM subregions are congruent with the lower similarity scores
seen in these regions.

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 3: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 3: Application to the segmentation of first episode
schizophrenia patients}

To validate that the \mb{} works effectively in the context of other neurological
disorders, in this experiment we use the Winterburn atlases to derive 
whole hippocampal segmentations of a dataset of patients with schizophrenia. The
resulting segmentations are assessed for quality by comparison with expert
manual segmentations.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%    RESULTS: Experiment 3, Application of \mb{} to the segmentation  %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Results}

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 4: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 4: Application to the segmentation of Alzheimer's
disease patients}

To validate \mb{} segmentation quality with respect to other established
automated hippocampal segmentation methods, \mb{} was applied to large dataset
from the ADNI project and the resulting segmentations were compared to those
produced by \freesurfer{}, \fsl{}, \maper{}, as well as semi-automated
whole hippocampal segmentations (\snt{}) provided by ADNI.

@@

\paragraph{Experiment details} 
\mb{} was configured with an atlas library composed of the five Winterburn T1
atlases (described in Experiment 2), and size of template library,
registration method, and label fusion method were determined by the optimal
settings found in Experiment 1.  The \fsl{} segmentation method was used via
the \verb+run_first_all+ script according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}.  All images
in the \adnidataset{} dataset were segmented by both methods.

@@

\paragraph{Evaluation method}
As in Experiment 3, the \snt{} and Winterburn hippocampal segmentation protocols
differ in the neuroanatomical features delineated, and so we assessed \mb{} by
the degree of (Pearson) correlation of average hippocampal volume across
subjects. We also computed the correlation in hippocampal volume between
existing, established automated segmentation methods -- \fsl{}, \freesurfer{},
and \maper{}, and \snt{} semi-automated segmentations. Additionally, we evaluate
the volume-related fixed and proportional biases in all segmentation methods
using Bland-Altman plots \citep{Bland1986}.

@@

Based on the results from Experiment 1, in this experiment \mb{} was configured
with a template library of 21 randomly chosen subject images (7 from each
disease class) and majority vote label fusion.  The entire \adnidataset{} dataset
was segmented by \mb{}, and the resulting volumes compared with those obtained by
expert semi-automated segmentation (\snt{}), and three other automated segmentation
techniques: \maper{}, \freesurfer{}, and \fsl{}. Table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the total count of
segmentations available, including a count of those which have failed a quality
control inspection.  A total of \Sexpr{length(means.complete$RID)} images are
included in the following analysis, having met quality criteria and having
segmentations from every method.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
%  - fix figure 2 schematic language for the experiments (not 2a 2b). 
%  - ADNI-1 or ADNI1, standardize
%  - Don't use "Experiment X", instead use some short form description, and then
%  reference (Experiment X). 
%  - subfield or subregion -- choose one

@@

THe gold standard for neuroanatomical segmentation is manual delineation by an
expert human rater.  However, with the availability of increasingly large MRI
datasets, the time and expertise required for manual segmentation becomes
prohibitive \citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. This
effort is complicated by the fact that there is significant variation between
segmentation protocols with respect to specific anatomical boundaries of the
hippocampus \citep{Geuze2004} and this has led to efforts to create an unified
hippocampal segmentation protocol \citep{Jack2011,Boccardi2013,Boccardi2013a}.
In addition, there is controversy over the appropriate manual segmentation
protocol to use in a particular imaging study \citep{Nestor2012}. Thus, a
segmentation algorithm that can easily adapt to different manual segmentation
definitions would be of significant benefit to the neuroimaging community.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                            Experiments                              %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experiments}
\label{sec:experiments}

The following section describes experiments conducted to assess the
segmentation quality of the \mb{} algorithm:  
\begin{itemize}
\item Experiment 1 investigates the accuracy of \mb{} whole hippocampus
segmentation of aging and Alzhiemer's diseased subjects over a wide range of
parameter settings using a Monte Carlo cross-validation design. The results of
this experiment enable us to choose the parameter settings offering the best
performance for use in subsequent experiments.  
\item Experiment 2 consists of a similar cross-validation design to verify these
parameter settings on brain images of young, first episode Schizophrenia
patients. 
\item Experiment 3 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. 
\item Experiment 4 compares \mb{} segmentations with those of well-known
automated methods on the entire \adnidataset image dataset consisting of 
\Sexpr{length(means.complete$RID)} brain images of subjects diagnosed as
cognitively normal, having mild cognitive impairment, or Alzheimer's disease.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 1: Whole Hippocampus Cross-Validation            %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 1: Whole Hippocampus Segmentation --- Cross-Validation on 
Alzheimer's Diseased Subjects}
\label{exp:adni-xval}

% TODO: properly introduce
Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using images
from the ADNI dataset \citep{Jack2008}, and manual whole hippocampus
segmentations following the protocol laid out in \citep{Pruessner2000}.  This
form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb{} (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

% TODO: fix wording, relate to pruessner labels
Note, in the Supplementary Materials we have replicated this experiment using the
\snt{} semi-automated labels included with the ADNI dataset.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 2: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2: Whole Hippocampus Segmentation --- Cross-validation
on First Episode Schizophrenia Subjects}
\label{exp:fep-xval}
%TODO: update
To validate that the \mb{} works effectively in the context of other neurological
disorders, in this experiment we use the Winterburn atlases to derive 
whole hippocampal segmentations of a dataset of patients with schizophrenia. The
resulting segmentations are assessed for quality by comparison with expert
manual segmentations.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%    RESULTS: Experiment 2, Application of \mb{} to the segmentation  %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Results}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 3: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 3: Hippocampal Subfield Segmentation --- Cross-Validation
on the Winterburn Atlases}
\label{exp:winterburn-xval}
The previous experiment assesses \mb{} peformance on whole hippocampus
segmentation. In this experiment, we evaluate \mb{} hippocampal subfield
segmentation of standard 3T T1-weighted images at $0.9m^3$-isotropic voxels.  We
use a modified leave-one-out cross-validation (LOOCV) design.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%    RESULTS: Experiment 3: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Results}

\begin{figure}
<<fig:WAxval-kappa>>=
wa.n.sim = read.csv('data/winterburn.xval.similarity.csv')
wa.n.sim.whole = read.csv('data/winterburn.xval.similarity.whole.csv')
wa.n.sim.whole$label = as.factor(wa.n.sim.whole$label)
levels(wa.n.sim.whole$label) <- list(
  "Whole"=1,
  "Whole"=2)
wa.n.sim.whole$res='0.9'
wa.n.sim.whole=wa.n.sim.whole[c('atlases','subject','res','label','k','se','sn','j')]

wa.n.sim$label = as.factor(wa.n.sim$label)
levels(wa.n.sim$label) <-list(
  "CA1"=1, "CA1"=101, 
  "CA2/CA3"=5,"CA2/CA3"=105, 
  "CA4/DG"=4,"CA4/DG"=104, 
  "Subiculum"=2,"Subiculum"=102, 
  "SR/SL/SM"=6,"SR/SL/SM"=106)

wa.n.sim = rbind(wa.n.sim, wa.n.sim.whole)

# kappa vs. region
ggplot(wa.n.sim, aes(x=as.factor(label),y=k)) + 
  geom_boxplot(fill='lightblue') + 
  scale_y_continuous(breaks=seq(0,1,0.1)) +
  xlab("Subregion") + 
  ylab("Dice's Similarity Coefficient (DSC)")

CA1   = subset(wa.n.sim,label=='CA1')$k
CA23  = subset(wa.n.sim,label=='CA2/CA3')$k
CA4DG = subset(wa.n.sim,label=='CA4/DG')$k
Sub   = subset(wa.n.sim,label=='Subiculum')$k
SRLM  = subset(wa.n.sim,label=='SR/SL/SM')$k
Whole = subset(wa.n.sim,label=='Whole')$k
@
  \caption{Similarity of \mb{} segmentation of subregions and the resampled
  Winterburn atlas segmentations at $0.9mm^3$ voxel resolution, over all folds of a
  leave-one-out cross validation. Overlap score for each hemisphere is measured
  separately. \label{fig:WAxval-kappa}}
\end{figure}

Figure \ref{fig:WAxval-kappa} shows the overlap similarity scores between the
\mb{} segmentations and the resampled Winterburn atlases for each hippocampal
subregion across all subjects and folds of the validation. Mean and standard
deviation DSC scores of the subregions are shown in Table
\ref{tab:WAxval-perturb-kappa}, along with DSC scores for the resampled atlas
segmentations when perturbed slightly and compared to the originals.  We find
that the CA4/DG subregion shows the highest mean DSC score of
\Sexpr{mean(CA4DG)} $\pm$\Sexpr{sd(CA4DG)}, followed by the Subiculum and CA1
subregions having scores of \Sexpr{mean(CA1)} $\pm$\Sexpr{sd(CA1)} and
\Sexpr{mean(Sub)} $\pm$\Sexpr{sd(Sub)}, respectively.  Both the CA4/DG and
molecular regions score below 0.5.  These scores may seem low but not when taken
in context and compared to existing (semi-)automated methods (see Discussion). The whole hippocampus is segmented with a mean DSC score of\Sexpr{mean(Whole)} $\pm$\Sexpr{sd(Whole)}.

\begin{figure}
<<fig:WAxval-Bland-Altman,fig.height=5,fig.width=7>>=
wa.n.vol = read.csv('data/winterburn.xval.volumes.csv') # xval volumes
winterburn = read.csv('data/winterburn.volumes.csv')    # resampled volumes
subfield_labels <-list(
  "CA1"="X1", "CA1"="X101", 
  "CA2/CA3"="X5","CA2/CA3"="X105", 
  "CA4/DG"="X4","CA4/DG"="X104",
  "Subiculum"="X2","Subiculum"="X102",
  "SR/SL/SM"="X6","SR/SL/SM"="X106", 
  "Whole" = "left", "Whole" = "right")
hemispheres <- list(
  "Right"="X1","Left"="X101", 
  "Right"="X5","Left"="X105", 
  "Right"="X4","Left"="X104",
  "Right"="X2","Left"="X102",
  "Right"="X6","Left"="X106",
  "Right"="right","Left"="left")

# compute whole hippocampal volumes (we do this the stupid way)
wa.n.vol = within(wa.n.vol, {
  right = X1+X2+X4+X5+X6
  left  = X101+X102+X104+X105+X106
  })
winterburn = within(winterburn, {
  right = X1+X2+X4+X5+X6
  left  = X101+X102+X104+X105+X106
  })

# mean per subfield/hemisphere across all 4 validation folds
wa.n.mean.vols = ddply(wa.n.all.vols, c("variable", "subject", "res"), function(df) {
  data.frame(mb    = mean(df$value),
             high  = mean(df$'0.3'),
             low   = mean(df$'0.9'))
})
wa.n.mean.vols$diff = wa.n.mean.vols$low - wa.n.mean.vols$mb
wa.n.mean.vols$mean = (wa.n.mean.vols$low + wa.n.mean.vols$mb)/2
wa.n.mean.vols$hemisphere        <- wa.n.mean.vols$variable
levels(wa.n.mean.vols$variable)  <- subfield_labels

wa.n.all.vols$diff = wa.n.all.vols$'0.9' - wa.n.all.vols$value
wa.n.all.vols$mean = (wa.n.all.vols$'0.9' + wa.n.all.vols$value)/2
wa.n.all.vols$Hemisphere        <- wa.n.all.vols$variable
levels(wa.n.all.vols$variable)   <- subfield_labels
levels(wa.n.all.vols$Hemisphere) <- hemispheres

ggplot(wa.n.all.vols, aes(x=mean, y=diff)) + 
  facet_wrap(~variable, ncol=3, scales='free_x') + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=2) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=wa.n.xval.bland.limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", 
             data = wa.n.xval.bland.limits, hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of resampled and MAGeT-Brain subregion volume ' (mm^3))) + 
  ylab(expression('Difference of manual and MAGeT-Brain subregion volume ' (mm^3))) + 
  #scale_x_continuous(limits=c(0,1000)) + 
  #scale_color_discrete(name="") + 
  scale_shape_discrete(name='Subject') + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@
  \caption{\textbf{Bland-Altman plots comparing unilateral hippocampal subregion volume
across all folds of Winterburn atlas cross validation.} The overall mean
difference in volume, and limits of agreement ($\pm 1.96SD$) are shown by dashed
horizontal lines. Linear fit lines are shown. Note, points below the mean
difference indication overestimation of the volume with respect to the resampled
volume, and vice versa. \label{fig:WAxval-Bland-Altman}}
\end{figure}

Figure \ref{fig:WAxval-Bland-Altman} contains Bland-Altman plots comparing \mb{}
volumes with manual volumes across all validation folds.  \mb{} displays a
conservative proportional bias --- small hippocampi are overestimated in volume,
and larger hippocampi are underestimated 
(a mean maximum difference of approximately 200mm^3 across all subfields). 
\mb{} display a slight conservative fixed bias, tending to underestimate all  
subfields except CA4/DG (mean underestimation: $CA1=76mm^3$, $CA2/3=56mm^3$,
$CA4/DG=-16mm^3$, $Subiculum=48mm^3$, $SR/SL/SM=96mm^3$).

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 4: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 4: Application to the segmentation of Alzheimer's
disease patients}
\label{exp:adni}
To validate \mb{} segmentation quality with respect to other established
automated hippocampal segmentation methods, \mb{} was applied to large dataset
from the ADNI project and the resulting segmentations were compared to those
produced by \freesurfer{}, \fsl{}, \maper{}, as well as semi-automated
whole hippocampal segmentations (\snt{}) provided by ADNI.

@@

\paragraph{Experiment details} 
\mb{} was configured with an atlas library composed of the five Winterburn T1
atlases (described in Experiment 3, section \ref{exp:winterburn-xval}), and size of template library,
registration method, and label fusion method were determined by the optimal
settings found in Experiment 1 (Section \ref{exp:adni-xval}).  The \fsl{} segmentation method was used via
the \verb+run_first_all+ script according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}.  All images
in the \adnidataset{} dataset were segmented by both methods.

@@

\paragraph{Evaluation method}
As in previous experiments, the \snt{} and Winterburn hippocampal segmentation protocols
differ in the neuroanatomical features delineated, and so we assess \mb{} by
the degree of (Pearson) correlation of average hippocampal volume across
subjects. We also computed the correlation in hippocampal volume between
existing, established automated segmentation methods -- \fsl{}, \freesurfer{},
and \maper{}, and \snt{} semi-automated segmentations. Additionally, we evaluate
the volume-related fixed and proportional biases in all segmentation methods
using Bland-Altman plots \citep{Bland1986}.

@@

Based on the results from Experiment 1 (Section \ref{exp:adni-xval}), in this experiment \mb{} was configured
with a template library of 21 randomly chosen subject images (7 from each
disease class) and majority vote label fusion.  The entire \adnidataset{} dataset
was segmented by \mb{}, and the resulting volumes compared with those obtained by
expert semi-automated segmentation (\snt{}), and three other automated segmentation
techniques: \maper{}, \freesurfer{}, and \fsl{}. Table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the total count of
segmentations available, including a count of those which have failed a quality
control inspection.  A total of \Sexpr{length(means.complete$RID)} images are
included in the following analysis, having met quality criteria and having
segmentations from every method.
